---
title: "MODUL FMI survey"
author: "Alex Abbas"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    fig_caption: yes
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, message = F, warning = F)
library(tidyverse)
library(readxl)
library(ComplexHeatmap)
library(DT)
library(knitr)
library(gridExtra)
library(cowplot)
library(survival)
library(survminer)


#truColors = read.csv("/Volumes/GoogleDrive/My Drive/R/trubetskoyColors.csv",stringsAsFactors = F)
# setwd("/Volumes/GoogleDrive/Shared drives/GI Biomarkers [gRED OBD]/CRC/MODUL Cohort 1/results/")
theme_set(theme_bw())

source("../R/loadData220218.R")
oapl = oaplv
```

# Introduction

MODUL (MO29112, NCT02291289) is an adaptable, phase 2, signal-seeking trial testing novel agents as first-line therapy for predefined subgroups of patients with metastatic colorectal cancer. It is comprised of four distinct cohorts. This analysis focuses on cohort 1, comprised of subjects with BRAF V600E mutant tumors, randomized to either the experimental arm of 5-FU/LV + cetuximab + vemurafenib or a control arm of FP + bevacizumab. Subjects' tumor tissue was tested with the FMI Foundation One CDx assay, and plasma samples from C1D1 and/or PD/EOT were tested with the FMI Foundation One Liquid CDx assay. Variants detected at these three points will be compared/contrasted to test hypotheses that treatment affected tumor genotype, specifically that cetuximab put a negative selective pressure on V600E mutations and a positive selective pressure on compensatory mutations elsewhere in the EGFR pathway. 

# Data

## Samples

```{r}
ospl %>% select(Visit) %>% group_by(Visit) %>% summarise(n = n()) %>% datatable()
```

## Foundation One CDx (i.e. Tissue Assay)

Survey June 15 2020 F1CDx tissue data, from OBD-FS data file:

    mo29112_ngs_dna_targdna_foundationone_20-2041_n65_one-alteration-per-line_20200615

```{r}
fmiGenes <- tissueOapl %>% select(GENE) %>% unique() %>% unlist()
tissueSubjs <- unique(tissueOapl$SUBJECT.ID)
```

Variants at baseline tissue biopsy:

```{r, fig.height=6, fig.width=10}


topGenes <- names(sort(table(tissueOapl$GENE),decreasing = T)[1:20])


```

# Results

## Variants at Biopsy and Cycle 1 Induction Phase

Compare gene-level variants for each patient between tissue and baseline (C1D1IP).

```{r, fig.height=7, fig.width=8}
genesAllTwo <- unique(unlist(
  liqOapl %>% filter(Visit == "C1D1IP") %>% select(GENE),
  tissueOapl %>% filter(Visit == "Tissue") %>% select(GENE)
))
topGenes <- names(sort(table(tissueOapl$GENE),decreasing = T)[1:50])
topGenesAllTwo <- intersect(topGenes,genesAllTwo)

tissC1Subjs = as.character(unique(intersect(
  liqOapl %>% filter(Visit == "C1D1IP") %>% select(SUBJECT.ID) %>% unlist(),
  tissueOapl %>% filter(Visit == "Tissue") %>% select(SUBJECT.ID) %>% unlist()
)))

twoMat = matrix(0,ncol=length(tissC1Subjs),nrow=length(topGenesAllTwo))
rownames(twoMat) = topGenesAllTwo
colnames(twoMat) = tissC1Subjs

tissueMut = tissueOapl %>% filter(SUBJECT.ID %in% tissC1Subjs & GENE %in% topGenesAllTwo) %>% select(SUBJECT.ID,GENE) %>% unique()
x = apply(tissueMut,1,function(oa){ twoMat[oa["GENE"],oa["SUBJECT.ID"]] <<- twoMat[oa["GENE"],oa["SUBJECT.ID"]] + 1 })

c1Mut = liqOapl %>% filter(Visit == "C1D1IP" & GENE %in% topGenesAllTwo) %>% select(SUBJECT.ID,GENE) %>% filter(SUBJECT.ID %in% tissC1Subjs) %>% unique()
x = apply(c1Mut,1,function(oa){ twoMat[oa["GENE"],oa["SUBJECT.ID"]] <<- twoMat[oa["GENE"],oa["SUBJECT.ID"]] + 2 })

class(twoMat) = "character"
mutHash = c(
  "0" = "WT",
  "1" = "Tissue only",
  "2" = "C1D1 only",
  "3" = "C1D1 & Tissue",
  "4" = "All"
)

mutCol = c(
  "WT" = "white",
  "Tissue only" = "red",
  "C1D1 only" = "blue",
  "C1D1 & Tissue" = "purple",
  "All" = "black"
)

mat2 = mutHash[twoMat]
dim(mat2) = dim(twoMat)
rownames(mat2) = rownames(twoMat)
colnames(mat2) = colnames(twoMat)
mat2 = mat2[,sort(colnames(mat2))]
# baseC1heatmap = Heatmap(mat2, col = mutCol, heatmap_legend_param = list(title = ""))
# baseC1heatmap
# pdf("../figures/baseToC1d1Oncoprint.pdf", height=4, width=8)
# baseC1heatmap
# dev.off()

oncoColsVs <- c(
  "WT" = "white",
  "Tissue only" = "red",
  "C1D1 only" = "blue",
  "C1D1 & Tissue" = "purple",
  "All" = "black",
  background = "white"
)

alter_funVs <- lapply(names(oncoColsVs),function(vartype) { function(x, y, w, h) grid.rect(x, y, w, h, gp = gpar(fill = oncoColsVs[vartype], col = NA)) })
names(alter_funVs) = names(oncoColsVs)
mat2[mat2=="WT"] = NA
baseC1heatmap = oncoPrint(mat2,
                          alter_fun = alter_funVs,
                          col = oncoColsVs,
                          top_annotation = NULL,
                          right_annotation = NULL,
                          show_pct = F,
                          show_column_names = T
)
baseC1heatmap
pdf("../figures/baseToC1d1Oncoprint220825b.pdf", height=5, width=8)
baseC1heatmap
dev.off()
  
```

## MAPK Pathway Genes

Focus on EGFR, KRAS, NRAS, BRAF, MAP2K1, NF1. Exclude V600E mutations here because they're present in all patients in this cohort. Plot presence/absence of each variant detected in these genes.

```{r, fig.height=8, fig.width=7}
#pathwayGenes = c("EGFR", "KRAS", "NRAS", "BRAF", "MAP2K1", "NF1")
#eot2pd = function(x) { x[x=="EOT"] = "PD"; x }

oaplPathway = oapl %>% filter(GENE %in% pathwayGenes & SV.PROTEIN.CHANGE != "V600E") %>%
  group_by(Arm,SUBJECT.ID,Response,GENE,Visit,variant) %>% summarise(SV.PERCENT.READS = sum(as.numeric(SV.PERCENT.READS),na.rm=T)) %>% ungroup() %>% 
#  mutate(Visit = eot2pd(Visit)) %>% 
  filter(Arm != "Control")

#visits = c("Tissue","C1D1IP","PD")
visits = c("Tissue","C1D1IP","induction","C1D1MP","ontx","PD","EOT") #unique(oapl$Visit)
#visits = c("Tissue","C1D1IP","C1D1MP","ontx","PD","EOT") #unique(oapl$Visit)
#visits = c("Tissue","C1D1IP","C1D1MP","ontx","PD")#,"EOT") #unique(oapl$Visit)

oaplPathwayWide = oaplPathway %>%
  filter(Visit %in% visits) %>%
  pivot_wider(names_from="Visit",values_from="SV.PERCENT.READS") %>%
  group_by(Arm, SUBJECT.ID, variant) %>% slice(n=1) %>% ungroup() %>% 
  arrange(GENE,Arm,SUBJECT.ID)

rowAnnotDf = data.frame(oaplPathwayWide %>% select(Arm))

annotCols = list(
  GENE = setNames(rainbow(length(intersect(unique(rowAnnotDf$GENE),pathwayGenes))),unique(rowAnnotDf$GENE)),
  Arm = setNames(c("darkslateblue","cadetblue2"),unique(rowAnnotDf$Arm))
#  Response = setNames(c("#55DD44","#1177AA","#CC4455","#AA55AA"),c("CR","PR","PD","SD"))
)

leftAnnot = rowAnnotation(
  df = rowAnnotDf,
  col = annotCols
)

toHeatmap = as.matrix(oaplPathwayWide[,visits])
class(toHeatmap) = "character"
rownames(toHeatmap) = paste(oaplPathwayWide$SUBJECT.ID,oaplPathwayWide$GENE,oaplPathwayWide$variant)
heatmapPatients = oaplPathwayWide$SUBJECT.ID
toHeatmap[!is.na(toHeatmap)] = "Mut"
toHeatmap[is.na(toHeatmap)] = "WT"

patientVisits = unlist(ospl %>% mutate(patientVisits = paste(SUBJECT.ID,Visit)) %>% select(patientVisits))
for (visit in visits) {
  for (i in 1:dim(toHeatmap)[1]) {
    patient <- heatmapPatients[i]
    rowname <- rownames(toHeatmap)[i]
    #    print(paste(rowname,patient,visit,paste(patient,visit) %in% patientVisits))
    if (!(paste(patient,visit) %in% patientVisits)) {
      toHeatmap[rowname,visit] <- "no data"
 #     print(paste(rowname,patient,visit,paste(patient,visit) %in% patientVisits))
    }
  }
}
rownames(toHeatmap) = paste(oaplPathwayWide$SUBJECT.ID,oaplPathwayWide$variant)

combineCols = function(aHeatmap,col1,col2,newName = "Final plasma") {
  newCol = aHeatmap[,col1]
  newCol[aHeatmap[,col2]=="Mut"] = "Mut"
  newCol[aHeatmap[,col2]=="WT" & newCol=="no data"] = "WT"
  aHeatmap = aHeatmap[,!colnames(aHeatmap) %in% c(col1,col2)]
  aHeatmap = cbind(aHeatmap,newCol)
  colnames(aHeatmap)[dim(aHeatmap)[2]] = newName
  aHeatmap
}

toHeatmap = combineCols(toHeatmap,"C1D1IP","induction",newName="Baseline plasma") # combine EOT and PD. Call Mut if it's Mut in either.
toHeatmap = combineCols(toHeatmap,"EOT","PD") 
toHeatmap = combineCols(toHeatmap,"C1D1MP","Final plasma")
toHeatmap = combineCols(toHeatmap,"ontx","Final plasma") 
toHeatmap = toHeatmap[,!colnames(toHeatmap) %in% c("C1D1MP")] # remove the C1D1MP timepoint



```



## Panel-wide screen

Our hypothesis focused on selected genes, but an unbiased screen of all genes on the FMI CDx panel would yield insight into whether this effect is significant because other genes could serve as a control.

```{r}
C1PdSubjs = as.character(intersect(
  oapl %>% filter(baseEnd == "base") %>% select(SUBJECT.ID) %>% unlist(),
  oapl %>% filter(baseEnd == "end") %>% select(SUBJECT.ID) %>% unlist()
))
```

Screen among the `r length(C1PdSubjs)` patients assessed at both C1D1 and PD/EOT for genes that show more than two known or likely impactful mutation gain and/or loss events.

```{r, fig.height=5, fig.width=5}
coh99toexp = function(x) { x[x=="Coh99"] = "Experimental"; x }
C1PdSubjDat = oapl %>%
  mutate(Arm = coh99toexp(Arm)) %>% 
  filter(SUBJECT.ID %in% C1PdSubjs) %>%
  mutate(variant = gsub("-","",paste0(SV.PROTEIN.CHANGE,CNA.TYPE,REARR.DESCRIPTION))) %>%
  # group_by(SUBJECT.ID,GENE,variant,baseEnd,Arm) %>% summarise(variants = toString(variant)) %>% ungroup() %>%
  select(SUBJECT.ID,GENE,variant,baseEnd,Arm,Tx) %>% unique() %>% 
  group_by(SUBJECT.ID,GENE,variant,Arm,Tx) %>% summarise(baseEnd = toString(baseEnd)) %>% ungroup() %>%
  filter(!grepl(",",baseEnd)) %>% 
  mutate(gainloss = ifelse(baseEnd == "base","loss","gain")) %>% unique()
C1PdSubjSum = C1PdSubjDat %>%
  group_by(GENE,gainloss,Arm,Tx) %>% group_by(GENE) %>% filter(n() > 2) %>% ungroup() %>%
  group_by(GENE,gainloss,Arm,Tx) %>% summarise(count = n()) %>% 
  group_by(Tx) %>% mutate(incidence = count/n())

C1PdSubjSum$count[C1PdSubjSum$gainloss == "loss"] = -C1PdSubjSum$count[C1PdSubjSum$gainloss == "loss"]
C1PdSubjSum$incidence[C1PdSubjSum$gainloss == "loss"] = -C1PdSubjSum$incidence[C1PdSubjSum$gainloss == "loss"]

geneOrder = C1PdSubjSum %>% group_by(GENE) %>% summarise(net = sum(count)) %>% ungroup() %>% arrange(net) %>% select(GENE) %>% unlist()

```


## Alteration Consistency

Survey how similar the detection of alterations and measurements of variant allele frequencies are between paired measurements.

```{r}
oapl[oapl=="-"] = ""
oapl$SV.PERCENT.READS[oapl$SV.PERCENT.READS==""] = "0"
oapl$SV.PERCENT.READS = as.numeric(oapl$SV.PERCENT.READS)


```

```{r}
tmb = tmb %>% mutate(
  Visit = factor(Visit,levels = c("Tissue","C1D1IP","C1D1MP","ontx","PD","EOT"))
)
```

## Oncoprint

```{r, fig.width = 12, fig.height = 4.5}
subjs <- unique(ospl$SUBJECT.ID)
subjs = subjs[order(as.numeric(gsub("\\D","",subjs)))]

# initialize the matrix with samples tested. Call them all wildtype first.
makeVisitMat = function(avisit) {
  # Make a full data matrix with everything initialized as wildtype
  mat <- matrix("",nrow=length(pathwayGenes),ncol=length(subjs))
  dimnames(mat) <- list(pathwayGenes,subjs)
  for( subjectid in unique( ospl %>% filter( Visit %in% avisit ) %>% select( SUBJECT.ID )) ) {
    mat[pathwayGenes,subjectid] = "wildtype"
  }
  
  # change from wildtype to V600E based on oapl
  pathwayOaplv600e <- oapl %>% filter(GENE %in% pathwayGenes & Visit %in% avisit & variant == "V600E")
  x <- apply(pathwayOaplv600e,1,function(oa){
    mat[oa["GENE"],oa["SUBJECT.ID"]] <<- "V600E"
  })
  
  # change from wildtype to other variant based on oapl
  pathwayOapl <- oapl %>%
    filter(GENE %in% pathwayGenes & Visit %in% avisit & variant != "V600E") %>% 
    arrange(desc(VARIANT.TYPE))
  x <- apply(pathwayOapl,1,function(oa){
    mat[oa["GENE"],oa["SUBJECT.ID"]] <<- oa["VARIANT.TYPE"]
  })
  
  # handle multiple variants
  pathway_multiSamples <- unlist(oapl %>%
    filter(GENE %in% pathwayGenes & Visit %in% avisit & variant != "V600E") %>% 
    select(GENE,SUBJECT.ID,Visit,VARIANT.TYPE,SAMPLE.ID) %>%
    unique() %>% 
    group_by(GENE,SUBJECT.ID,Visit,SAMPLE.ID) %>% 
    summarise(n = n()) %>% filter(n == 2) %>%
    ungroup() %>% 
    select(SAMPLE.ID))
  pathwayOapl_multi = oapl %>% 
    filter(GENE %in% pathwayGenes & Visit %in% avisit & variant != "V600E") %>% 
    filter(SAMPLE.ID %in% pathway_multiSamples)
  x <- apply(pathwayOapl_multi,1,function(oa){
    mat[oa["GENE"],oa["SUBJECT.ID"]] <<- "multiple"
  })

  mat
}
matTissue = makeVisitMat("Tissue")
matC1 = makeVisitMat(c("C1D1IP","induction"))
matPD = makeVisitMat(c("C1D1MP","ontx","EOT","PD"))
mat = rbind(matTissue,matC1,matPD)
c1PdRowSplit = factor(
  c(
    rep("Archival\nTissue",times = dim(matTissue)[1]),
    rep("Baseline\nPlasma",times = dim(matC1)[1]),
    rep("Final\nPlasma",times = dim(matPD)[1])
  ),
  levels = c(
    "Archival\nTissue",
    "Baseline\nPlasma",
    "Final\nPlasma"
  )
)

responseCols <- c(
  "CR" = "#22CC00",
  "PR" = "#DDDD00",
  "SD" = "#DD8811",
  "PD" = "#CC3366"
)

tissueTmbDat = tmb %>% filter(Visit == "Tissue")
tissueTmbDat = tissueTmbDat[match(colnames(mat),tissueTmbDat$SUBJECT.ID),]
patientHeatDat <- patientData[match(colnames(mat),patientData$SUBJECT.ID),] %>% 
  mutate(
    tissueTMB = as.numeric(tissueTmbDat$TMB.SCORE),
    Response = factor(Response,levels = names(responseCols))
  )
subjSplit = patientHeatDat %>% select(SUBJECT.ID,"Tx") %>% unique()
splitFactor = factor(
  unlist(subjSplit[match(subjs,subjSplit$SUBJECT.ID),"Tx"]),
  levels = c("Control","Experimental")
)

oncoCols <- c(
  "short-variant" = "forestgreen",
  "V600E" = "#BBCCBB",
  "copy-number-alteration" = "darkorange",
  rearrangement = "purple",
  multiple = "magenta",
  wildtype = "gray90",
  background = "white"
)

alterFun <- lapply(names(oncoCols),function(vartype) { function(x, y, w, h) grid.rect(x, y, w, h, gp = gpar(fill = oncoCols[vartype], col = NA)) })
names(alterFun) = names(oncoCols)

annotCols = list(BOR = responseCols)


```

### with annotation marks for acquired variants

```{r, fig.height=7, fig.width=9}
leftAnnot = HeatmapAnnotation(
  "TMB\n(Mut/MB)" = anno_barplot(patientHeatDat$tissueTMB, border = F, axis_param = c(direction="reverse")),
  BOR = patientHeatDat$Response,
  col = annotCols,
  which = "row"
)

op2 = oncoPrint(t(mat),
          alter_fun = alterFun,
          col = oncoCols,
          show_pct = F,
          row_split = splitFactor,
          column_split = c1PdRowSplit,
          left_annotation = leftAnnot,
          right_annotation = NULL,
          top_annotation = NULL,
          row_order = order(patientHeatDat$Response),
          show_row_names = T,
          show_column_names = T,
          row_names_gp = gpar(fontsize = 9)
 )

preTimes = c("Tissue","C1D1IP","induction")
postTimes = c("C1D1MP","ontx","EOT","PD")

oapl = oapl %>% mutate(subjectGeneVariant = paste(SUBJECT.ID,GENE,variant))

preSGVs = (oapl %>% filter(Visit %in% preTimes))$subjectGeneVariant
acqPathVars = oapl %>%
  filter(
    GENE %in% pathwayGenes &
      variant != "V600E" &
      Visit %in% postTimes &
      !( subjectGeneVariant %in% preSGVs)
  ) %>% 
  group_by(SUBJECT.ID,GENE) %>% summarise(variants = paste(unique(variant), collapse = " ")) %>%
  ungroup() %>% 
  mutate(geneVariants = paste(GENE,variants)) %>% 
  group_by(SUBJECT.ID) %>% summarise(geneVariants = paste(geneVariants, collapse = ", ")) %>%
  ungroup() %>% 
  select(SUBJECT.ID,geneVariants) %>% unique()

labels_list = unlist(lapply(acqPathVars$geneVariants, function(x) 
  paste0(strwrap(paste0(x, collapse = " "), width = 45), collapse = "\n")))

variantMarks = anno_mark(
  at = match(acqPathVars$SUBJECT.ID, colnames(mat)),
  labels = labels_list,
  which = "row",
  padding = unit(3, "mm"),
  labels_gp = gpar(fontsize = 10)
)

op2mark = op2 + rowAnnotation(mark = variantMarks)
op2mark
# pdf("/Volumes/GoogleDrive/Shared drives/GI Biomarkers [gRED OBD]/CRC/MODUL Cohort 1/figures/C1_PD_oncoprint_220426txMark.pdf", width=10, height=8)
# op2mark
# dev.off()
```

### with annotation marks for acquired variants, split by arm

```{r, fig.height=7, fig.width=9}
leftAnnot = HeatmapAnnotation(
  "TMB\n(Mut/MB)" = anno_barplot(patientHeatDat$tissueTMB, border = F, axis_param = c(direction="reverse")),
  BOR = patientHeatDat$Response,
  col = annotCols,
  which = "row"
)

subjSplit = patientHeatDat %>% select(SUBJECT.ID,"Arm") %>% unique()
splitFactor = factor(
  unlist(subjSplit[match(subjs,subjSplit$SUBJECT.ID),"Arm"]),
  levels = c("Control","Experimental","Coh99")
)

op2 = oncoPrint(t(mat),
          alter_fun = alterFun,
          col = oncoCols,
          show_pct = F,
          row_split = splitFactor,
          column_split = c1PdRowSplit,
          left_annotation = leftAnnot,
          right_annotation = NULL,
          top_annotation = NULL,
          row_order = order(patientHeatDat$Response),
          show_row_names = T,
          show_column_names = T,
          row_names_gp = gpar(fontsize = 9)
 )

preTimes = c("Tissue","C1D1IP","induction")
postTimes = c("C1D1MP","ontx","EOT","PD")

oapl = oapl %>% mutate(subjectGeneVariant = paste(SUBJECT.ID,GENE,variant))

preSGVs = (oapl %>% filter(Visit %in% preTimes))$subjectGeneVariant
acqPathVars = oapl %>%
  filter(
    GENE %in% pathwayGenes &
      variant != "V600E" &
      Visit %in% postTimes &
      !( subjectGeneVariant %in% preSGVs)
  ) %>% 
  group_by(SUBJECT.ID,GENE) %>% summarise(variants = paste(unique(variant), collapse = " ")) %>%
  ungroup() %>% 
  mutate(geneVariants = paste(GENE,variants)) %>% 
  group_by(SUBJECT.ID) %>% summarise(geneVariants = paste(geneVariants, collapse = ", ")) %>%
  ungroup() %>% 
  select(SUBJECT.ID,geneVariants) %>% unique()

labels_list = unlist(lapply(acqPathVars$geneVariants, function(x) 
  paste0(strwrap(paste0(x, collapse = " "), width = 45), collapse = "\n")))

variantMarks = anno_mark(
  at = match(acqPathVars$SUBJECT.ID, colnames(mat)),
  labels = labels_list,
  which = "row",
  padding = unit(3, "mm"),
  labels_gp = gpar(fontsize = 10)
)

op2mark = op2 + rowAnnotation(mark = variantMarks)
op2mark
# pdf("/Volumes/GoogleDrive/Shared drives/GI Biomarkers [gRED OBD]/CRC/MODUL Cohort 1/figures/C1_PD_oncoprint_220426armMark.pdf", width=10, height=8)
# op2mark
# dev.off()
```

### Plots of patient-level MAPK variant incidence

```{r}
treatMapkCounts = ospl %>%
  select(Arm,SUBJECT.ID) %>%
  unique() %>% 
  mutate(
    Treatment = ifelse(Arm=="Control","Control","Experimental"),
    MAPKvariant = ifelse(SUBJECT.ID %in% acqPathVars$SUBJECT.ID,"Variant Acquired","No change")) %>%
  select(Treatment,MAPKvariant) %>% 
  table() %>% data.frame() %>% 
  group_by(Treatment) %>%
  mutate(
    Freq2 = Freq,
    pct = round(Freq / sum(Freq)*100),
    myLabel = paste0(Freq,"\n(",pct,"%)")
  )

treatMapkCounts[treatMapkCounts$Freq==0,"Freq"] = NA
dots = ggplot(treatMapkCounts, aes(x = Treatment, y = MAPKvariant, size = Freq, label = myLabel)) +
  geom_point(col = "darkblue") + scale_size_area(max_size = 50) +
  ylab("MAPK Pathway Status") +
  theme(legend.position="none") +
  geom_text(col="white",aes(size=sqrt(Freq2)*.15))
dots
# pdf("/Volumes/GoogleDrive/Shared drives/GI Biomarkers [gRED OBD]/CRC/MODUL Cohort 1/figures/MAPKdots.pdf", width=4.5, height=4)
# dots
# dev.off()


# Donut plot
treatMapkCounts = treatMapkCounts %>% group_by(Treatment) %>%   # Compute the cumulative percentages (top of each rectangle)
  mutate(
    ymax = cumsum(Freq2),
    ymin = c(0, head(ymax, n=-1))
  )

ctrlDonut = ggplot(treatMapkCounts %>% filter(Treatment=="Control"), aes(ymax=ymax, ymin=ymin, xmax=3.5, xmin=3, fill=MAPKvariant)) +
     geom_rect() +
     coord_polar(theta="y") + # Try to remove that to understand how the chart is built initially
     xlim(c(2, 4)) + # Try to remove that to see how to make a pie chart
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank(),legend.position="none",
          panel.background=element_blank(),panel.border=element_blank(),panel.grid.major=element_blank(),
          panel.grid.minor=element_blank(),plot.background=element_blank())
expDonut = ggplot(treatMapkCounts %>% filter(Treatment=="Experimental"), aes(ymax=ymax, ymin=ymin, xmax=4, xmin=3, fill=MAPKvariant)) +
     geom_rect() +
     coord_polar(theta="y") + # Try to remove that to understand how the chart is built initially
     xlim(c(2, 4)) + # Try to remove that to see how to make a pie chart
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank(),legend.position="none",
          panel.background=element_blank(),panel.border=element_blank(),panel.grid.major=element_blank(),
          panel.grid.minor=element_blank(),plot.background=element_blank())

# pdf("/Volumes/GoogleDrive/Shared drives/GI Biomarkers [gRED OBD]/CRC/MODUL Cohort 1/figures/MAPKdonuts.pdf", width=8, height=4)
# plot_grid(ctrlDonut,expDonut)
# dev.off()
# ggplot(treatMapkCounts, aes(ymax=ymax, ymin=ymin, xmax=4, xmin=3, fill=MAPKvariant)) +
#      geom_rect() +
#      coord_polar(theta="y") + # Try to remove that to understand how the chart is built initially
#      xlim(c(2, 4)) + # Try to remove that to see how to make a pie chart
#   facet_wrap(~Treatment) +
#   theme(axis.line=element_blank(),axis.text.x=element_blank(),
#           axis.text.y=element_blank(),axis.ticks=element_blank(),
#           axis.title.x=element_blank(),
#           axis.title.y=element_blank(),legend.position="none",
#           panel.background=element_blank(),panel.border=element_blank(),panel.grid.major=element_blank(),
#           panel.grid.minor=element_blank(),plot.background=element_blank())
```


```{r}
subjSplit = patientHeatDat %>% select(SUBJECT.ID,"Arm") %>% unique()
splitFactor = factor(
  unlist(subjSplit[match(subjs,subjSplit$SUBJECT.ID),"Arm"]),
  levels = c("Control","Experimental","Coh99")
)

oncoCols <- c(
  "short-variant" = "forestgreen",
  "V600E" = "#BBCCBB",
  "copy-number-alteration" = "darkorange",
  rearrangement = "purple",
  multiple = "magenta",
  wildtype = "gray90",
  background = "white"
)

alterFun <- lapply(names(oncoCols),function(vartype) { function(x, y, w, h) grid.rect(x, y, w, h, gp = gpar(fill = oncoCols[vartype], col = NA)) })
names(alterFun) = names(oncoCols)

annotCols = list(BOR = responseCols)

topAnnot = HeatmapAnnotation(
  "TMB\n(Mut/MB)" = anno_barplot(patientHeatDat$tissueTMB,border = F),
  BOR = patientHeatDat$Response,
  col = annotCols
)

op1 = oncoPrint(mat,
          alter_fun = alterFun,
          col = oncoCols,
          show_pct = F,
          column_split = splitFactor,
          row_split = c1PdRowSplit,
          top_annotation = topAnnot,
          right_annotation = NULL,
          left_annotation = NULL,
          column_order = order(patientHeatDat$Response),
          show_column_names = T
 )

op1

# pdf("/Volumes/GoogleDrive/Shared drives/GI Biomarkers [gRED OBD]/CRC/MODUL Cohort 1/figures/C1_PD_oncoprint_220317arm.pdf", width=10, height=3.5)
# op1
# dev.off()
```

## Swimmer plot

```{r, fig.height=11, fig.width=8}
# make a data frame with one sample per row and presence/absence of detected MAPK pathway mutation
#nonBrafGenes <- c("EGFR", "KRAS", "NRAS", "MAP2K1", "NF1")
oapl = oapl %>% 
  mutate(
    mapkMut = GENE %in% pathwayGenes & variant != "V600E"
  )
osplPath = oapl %>%
  select(SUBJECT.ID,SAMPLE.ID,mapkMut,Cycle,Visit,timepoint,Tx) %>% 
  group_by(SUBJECT.ID,Cycle,Visit,timepoint,Tx,SAMPLE.ID) %>%
  summarise(state = ifelse(any(mapkMut),"Positive","Negative")) %>% 
  mutate(
    timepointMonth = timepoint / 52 * 12,
    timepointDay = timepoint / 52 * 365,
    Event = ifelse(Visit=="Tissue","Archival Biopsy","ctDNA")
  ) %>% 
  ungroup()

# fill in events with no mutations in oapl
toAdd = ospl %>%
  filter( !(SAMPLE.ID %in% osplPath$SAMPLE.ID) ) %>%
  select( SUBJECT.ID, Cycle, Visit, Tx, timepoint) %>% 
  dplyr::mutate(
    timepointMonth = timepoint / 52 * 12,
    timepointDay = timepoint / 52 * 365,
    Event = "ctDNA",
    state = "Negative"
  )
osplPath = osplPath %>% bind_rows(toAdd)

# add rows for submitted samples
# submittedSamples = read_excel("/Volumes/GoogleDrive/Shared drives/GI Biomarkers [gRED OBD]/CRC/MODUL Cohort 1/data/Extra FMI samples Modul cohort 1.xlsx")
# submittedSamples = submittedSamples %>% 
#   filter(SUBJECT.ID != "3207") %>% 
#   mutate(
#     SUBJECT.ID = as.character(SUBJECT.ID),
#     timepoint = cycleToTime(Cycle),
#     timepoint = timepoint / 52 * 12,
#     Event = "ctDNA",
#     state = "Submitted"
#   ) %>% left_join(patientData %>% select(SUBJECT.ID,Tx))
# osplPath = osplPath %>%
#   bind_rows(submittedSamples) %>% 
#   mutate(PostTx = state)

# add clinical timecourse response data
clinToBind = clinicalData %>%
  select(SUBJECT.ID,ADY,AVALC,APHASE) %>% 
  dplyr::rename(
    state = AVALC
  ) %>% 
  mutate(
    timepointMonth = as.numeric(ADY) / 365 * 12,
    timepointDay = as.numeric(ADY),
    Event = "Scan"
  ) %>% 
  filter(
    SUBJECT.ID %in% osplPath$SUBJECT.ID
  ) %>% 
  left_join(osplPath %>% select(SUBJECT.ID,Tx) %>% unique()) %>% 
  mutate(
    PostTx = ifelse(APHASE == "Post-trt" | Event == "ctDNA", "yes", "no")
  )
osplClin = bind_rows(clinToBind,osplPath)

# make a swimmer plot with it
strokeColors = c(
  "Negative" = "gray20",
  "Positive" = "gray20",
#  "Submitted" = "#4488FF",
  "yes" = "gray20",
  "no" = "#00000000",
  "CR" = "#22CC00",
  "PR" = "#DDDD00",
  "SD" = "#DD8811",
  "PD" = "#CC3366"
)

tdf = "triangle down filled"
tf = "triangle filled"
sq = "square filled"
cf = "circle filled"

eventPch = c(
  "Archival Biopsy" = cf,
  "ctDNA" = tf,
  "Scan" = sq
)

stateFills = c(
  "Negative" = "white",
  "Positive" = "gray20",
  "Submitted" = "#4488FF",
  "Complete Response (CR)" = "#22CC00",
  "Partial Response (PR)" = "#DDDD00",
  "Stable Disease (SD)" = "#DD8811",
  "Progressive Disease (PD)" = "#CC3366",
  "Not Evaluable (NE)" = "gray50"
)

osplClin$state = factor(osplClin$state, levels = names(stateFills))
osplClin = osplClin %>% left_join(patientData)

gp <- ggplot(osplClin, aes(x = timepointDay, y = SUBJECT.ID, fill = state, alpha = state, col = PostTx, shape = Event, group = SUBJECT.ID)) +
  geom_point(data = osplClin %>% filter(Event == "Scan"), size = 2.3, position = position_nudge( y = -.18 ) ) +
  geom_line(aes(color = Response), size = 1.5) +
  geom_point(data = osplClin %>% filter(Event == "ctDNA"), size = 1.9, position = position_nudge( y = .2 ) ) +
  geom_point(data = osplClin %>% filter(Event == "Archival Biopsy"), size = 2.5 ) +
  theme_classic() +
  facet_grid(rows = "Tx", space = "free", scales = "free", ) +
  xlab("Time relative to start of maintenance phase (days)") +
  scale_color_manual(
    values = strokeColors,
    breaks = c("yes","no"),
    guide = guide_legend(override.aes = list(shape = sq, linetype = 0, fill = "gray80"), title = "Post-tx")) +
  scale_fill_manual(
    values = stateFills,
    name = "RECIST",
    breaks = c(  "Complete Response (CR)", "Partial Response (PR)", "Stable Disease (SD)", "Progressive Disease (PD)", "Not Evaluable (NE)"),
    guide = guide_legend(override.aes = list(color = "gray50", linetype = 1, shape = sq) ) ) +
  scale_shape_manual(values = eventPch) +
  scale_alpha_manual(
    name = "MAPK pathway ctDNA status",
    values = c(1, 1, 1, 1, 1, 1, 1, 1),
    breaks = c("Positive", "Negative"),#, "Submitted"),
    guide = guide_legend(
      override.aes = list(
        linetype = 0,
        shape = tf,
        fill = c("gray20","white"),#,"#4488FF"),
        color = c("gray20","gray20")#,"#4488FF")
      ))) +
  theme(legend.position = c(0.85, 0.75)) +
  scale_x_continuous(breaks = c(1:6)*10-20, labels = c("Archival",c(1:5)*10-10))

# pdf("/Volumes/GoogleDrive/Shared drives/GI Biomarkers [gRED OBD]/CRC/MODUL Cohort 1/figures/swimmerplot_all_211109.pdf", width=8, height=12)
# print(gp)
# dev.off()
gp


```

Just for selected subjects.

```{r, fig.height = 5.5, fig.width = 8}
mapkSubjectids <- c(setdiff(
  unlist(osplClin %>% filter(Visit == "PD" & state == "Positive" & !is.na(timepointMonth)) %>% select(SUBJECT.ID)),
  unlist(osplClin %>% filter(Visit == "C1D1IP" & state == "Positive" & !is.na(timepointMonth)) %>% select(SUBJECT.ID))
), "E2", "E26", "E5", "E32")
osplClin2 <- osplClin %>% filter(SUBJECT.ID %in% mapkSubjectids & Tx == "Experimental")

gp <- ggplot(osplClin2, aes(x = timepointDay, y = SUBJECT.ID, fill = state, alpha = state, col = PostTx, shape = Event, group = SUBJECT.ID)) +
  geom_point(data = osplClin2 %>% filter(Event == "Scan"), size = 3, position = position_nudge( y = -.10 ) ) +
  geom_line(aes(color = Response), size = 1.5) +
  geom_point(data = osplClin2 %>% filter(Event == "ctDNA"), size = 2.5, position = position_nudge( y = .12 ) ) +
  geom_point(data = osplClin2 %>% filter(Event == "Archival Biopsy"), size = 3 ) +
  theme_classic() +
  theme( axis.text = element_text( size = 13 ),
         axis.title = element_text(size=13, face = "bold") ) +
  facet_grid(rows = "Tx", space = "free", scales = "free", ) +
  xlab("Time relative to start of maintenance phase (days)") +
  scale_color_manual(
    values = strokeColors,
    breaks = c("yes","no"),
    guide = guide_legend(override.aes = list(shape = sq, linetype = 0, fill = "gray80"), title = "Post-tx")) +
  scale_fill_manual(
    values = stateFills,
    name = "RECIST",
    breaks = c(  "Complete Response (CR)", "Partial Response (PR)", "Stable Disease (SD)", "Progressive Disease (PD)", "Not Evaluable (NE)"),
    guide = guide_legend(override.aes = list(color = "gray50", linetype = 1, shape = sq) ) ) +
  scale_shape_manual(values = eventPch) +
  scale_alpha_manual(
    name = "MAPK pathway ctDNA status",
    values = c(1, 1, 1, 1, 1, 1, 1, 1),
    breaks = c("Positive", "Negative", "Submitted"),
    guide = guide_legend(
      override.aes = list(
        linetype = 0,
        shape = tf,
        fill = c("gray20","white"),#,"#4488FF"),
        color = c("gray20","gray20")#,"#4488FF")
      ) ) ) +
  scale_x_continuous(breaks = c(1:6)*10-20, labels = c("Archival",c(1:5)*10-10))


# pdf("/Volumes/GoogleDrive/Shared drives/GI Biomarkers [gRED OBD]/CRC/MODUL Cohort 1/figures/swimmerplot_selected_211109.pdf", width=8, height=5.5)
# print(gp)
# dev.off()
gp
```

Ordered by last scan date, with genotype colors removed.

```{r, fig.height=11, fig.width=8}
subjInOrder = osplClin %>%
  filter(!is.na(timepointMonth)) %>%
  group_by(SUBJECT.ID) %>% summarise(timepointMonth = max(timepointMonth)) %>% ungroup() %>% arrange(desc(timepointMonth)) %>%
  select(SUBJECT.ID) %>% unlist()
osplClin3 = osplClin %>% mutate(
  SUBJECT.ID = factor(SUBJECT.ID, levels = subjInOrder)
)
stateFills = c(
  "Negative" = "gray50",
  "Positive" = "gray50",
  "Submitted" = "#4488FF",
  "Complete Response (CR)" = "#22CC00",
  "Partial Response (PR)" = "#DDDD00",
  "Stable Disease (SD)" = "#DD8811",
  "Progressive Disease (PD)" = "#CC3366",
  "Not Evaluable (NE)" = "gray50"
)

gp2 <- ggplot(osplClin3, aes(x = timepointMonth, y = SUBJECT.ID, fill = state, alpha = state, col = PostTx, shape = Event, group = SUBJECT.ID)) +
  geom_point(data = osplClin3 %>% filter(Event == "Scan"), size = 2, position = position_nudge( y = -.18 ) ) +
  geom_line(aes(color = Response), size = 1.5) +
  geom_point(data = osplClin3 %>% filter(Event == "ctDNA"), size = 2.3, position = position_nudge( y = .2 ) ) +
  geom_point(data = osplClin3 %>% filter(Event == "Archival Biopsy"), size = 2.5 ) +
  theme_classic() +
  facet_grid(rows = "Tx", space = "free", scales = "free", ) +
  xlab("Time relative to start of maintenance phase (months)") +
  scale_color_manual(
    values = strokeColors,
    breaks = c("yes","no"),
    guide = guide_legend(override.aes = list(shape = sq, linetype = 0, fill = "gray80"), title = "Post-tx")) +
  scale_fill_manual(
    values = stateFills,
    name = "RECIST",
    breaks = c(  "Complete Response (CR)", "Partial Response (PR)", "Stable Disease (SD)", "Progressive Disease (PD)", "Not Evaluable (NE)"),
    guide = guide_legend(override.aes = list(color = "gray50", linetype = 1, shape = sq) ) ) +
  scale_shape_manual(values = eventPch) +
  scale_alpha_manual(
    name = "MAPK pathway ctDNA status",
    values = c(1, 1, 1, 1, 1, 1, 1, 1),
    breaks = c("Positive", "Negative"),#, "Submitted"),
    guide = guide_legend(
      override.aes = list(
        linetype = 0,
        shape = tf,
        fill = c("gray20","white"),#,"#4488FF"),
        color = c("gray20","gray20")#,"#4488FF")
      ))) +
  theme(legend.position = "none") + #c(0.85, 0.75)) +
  scale_x_continuous(breaks = c(1:6)*10-20, labels = c("Archival",c(1:5)*10-10))

# pdf("/Volumes/GoogleDrive/Shared drives/GI Biomarkers [gRED OBD]/CRC/MODUL Cohort 1/figures/swimmerplot_gray_211109.pdf", width=8, height=12)
# print(gp2)
# dev.off()
gp2

```

## Survival Analysis

Test whether patients that acquire a MAPK pathway mutation tend to have shorter survival times than those who don't. Turns out there's a trend, but it's not significant.

```{r}
osplSurv = osplClin %>%
  mutate(ADY = pmax(ADY,timepoint * 7,na.rm=T)) %>% 
  group_by(SUBJECT.ID) %>% summarise(tte = max(ADY,na.rm=T)) %>% mutate(mapk = SUBJECT.ID %in% mapkSubjectids, event = 1) %>% filter(tte>0)

cox1 = coxph(Surv(tte,event) ~ mapk, data = osplSurv)
summary(cox1)

iformula <- as.formula("Surv(tte, event) ~ mapk")
fit <- surv_fit(iformula, data = osplSurv)
#  names(fit$strata) <- gsub(".+=", "", names(fit$strata))
ggsurvplot(
  fit,
  #    pval = coxStatText,#, conf.int = TRUE,
  risk.table = T, #risk.table.col = strataCols, # Add risk table
  risk.table.y.text.col = T,
  risk.table.y.text = F,
  # linetype = lty, # Change line type by groups
  surv.median.line = "hv", # Specify median survival
  # # ggtheme = theme_bw(), # Change ggplot2 theme
  #    ylab = ylab,
  xlab = "Time to event (days)",
  #    palette = strataCols,
  #    title = main,
  #    font.title = c(14, "bold"),
  #    legend.title=""
)

```

## Longitidinal Analysis

Look at relationships between when variants appear and when disease progresses. Consider the first non-V600E MAPK mutation of known or likely consequence to be a positive variant event, as in the swimmer plots. Consider the first scan date with a RECIST status of Progressive Disease to be a positive scan event. Patients without both events are excluded here.

First, a density and rug plot of the interval between positive events.

```{r}
firstEvent = osplClin %>%
  filter(state %in% c("Positive","Progressive Disease (PD)") & Event %in% c("ctDNA","Scan")) %>% 
  group_by(SUBJECT.ID,Event,state) %>% summarise(timepointMonth = min(timepointMonth)) %>% slice(n=1) %>% ungroup() %>%
  select(-state) %>% 
  pivot_wider(values_from = "timepointMonth", names_from = "Event") %>% 
  filter(!is.na(ctDNA) & !is.na(Scan)) %>% 
  mutate(mut2PD = Scan - ctDNA)
ggplot(firstEvent,aes(x=mut2PD)) +
  geom_density(adjust = 1/2) +
  xlab("Time from MAPK mutation to PD (months)") +
  geom_rug()
```

Second, a scatter plot of variant event vs. scan event. Dashed line is x=y reference.

```{r}
ggplot(firstEvent,aes(x=ctDNA,y=Scan)) +
  geom_point(size=2,alpha=.5) +
  geom_hline(yintercept=0) +
  geom_vline(xintercept=0) +
  geom_abline(lty = "dashed") +
  xlab("Time of first MAPK mutation (months)") +
  ylab("Time of first RECIST PD (months)")

medianM2P = signif(quantile(firstEvent$mut2PD,c(.25,.5,.75)),digits=2)
```
The median time from variant to PD is `r medianM2P[2]` months. (quartiles `r medianM2P[1]`, `r medianM2P[3]`)

```{r}
sessionInfo()
```
