---
title: "MODUL FMI Patient Facet"
author: "Alex Abbas"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, message = F, warning = F, echo = F)
library(tidyverse)
library(readxl)
library(ComplexHeatmap)
library(DT)
library(knitr)
library(ggvenn)
library(UpSetR)
#library(stringr) 
#library(ggpubr)

#source("/Volumes/GoogleDrive/My Drive/R/aaStuff.R")
#truColors = read.csv("/Volumes/GoogleDrive/My Drive/R/trubetskoyColors.csv",stringsAsFactors = F)

theme_set(theme_bw())

# setwd("/Volumes/GoogleDrive/Shared drives/GI Biomarkers [gRED OBD]/CRC/MODUL Cohort 1/vignettes/")

#  data_dir="/gne/data/obdroot/avastin/mo29112_modul/ngs/results/ctdna/data"
ctDnaDir <- "../data/ctDNA FMI/"
tissueDir <- "../data/FMI/"
Aug30fmiDir <- "/Volumes/OBD/avastin/mo29112_modul/ngs/rawdata/fmi_obd_mo29112_ngs_dna_targdna_foundationoneliquidcdx_20-2047_std_20210830/"

# patientData
patientData <- read_excel("../data/MO29112_Ch1 clinical-biomarker dashboard (2).xlsx")
patientData <- patientData %>%
  rename(Arm = ARM, Response = BOR, SUBJECT.ID = "Subject ID") %>%
  select(SUBJECT.ID,Arm,Response) %>% 
  mutate(SUBJECT.ID = as.character(SUBJECT.ID))

# ctDNA data
oapl_ctdna_File <- paste0(
  ifelse ( F & dir.exists(Aug30fmiDir), Aug30fmiDir, ctDnaDir ),
  "mo29112_ngs_dna_targdna_foundationoneliquidcdx_20-2047_n183_one-alteration-per-line-plus_20210830.txt"
)
liqOapl <- read.delim(oapl_ctdna_File,sep="\t")
ospl_ctdna_file = paste0(ctDnaDir,"mo29112_ngs_dna_targdna_foundationoneliquidcdx_20-2047_n183_one-sample-per-line_20210830.xlsx")
liqOspl <- read_excel(ospl_ctdna_file) %>%
  rename(FMI.SAMPLE.ID = "FMI SAMPLE ID", SAMPLE.ID = "SAMPLE ID", SUBJECT.ID = "SUBJECT ID") %>% 
  mutate(
    SAMPLE.ID = as.character(SAMPLE.ID),
    Visit = gsub("\\W","",Visit),
    Cycle = gsub("\\W","",Cycle),
    SUBJECT.ID = gsub("\\.0","",SUBJECT.ID)
  )

liqOapl <- liqOapl %>%
  mutate(SAMPLE.ID = as.character(SAMPLE.ID),SUBJECT.ID = as.character(SUBJECT.ID)) %>% 
  left_join(liqOspl %>% select(FMI.SAMPLE.ID, Cycle, Visit))
# Tissue data
oapl_tissue_file <- paste0(
  ifelse ( F & dir.exists(Aug30fmiDir), Aug30fmiDir, tissueDir ),
  "mo29112_ngs_dna_targdna_foundationone_20-2041_n65_one-alteration-per-line_20200615.txt"
)
tissueOapl = read.delim(oapl_tissue_file,sep="\t") %>% 
  mutate(SUBJECT.ID = as.character(SUBJECT.ID))
ospl_tissue_file <- paste0(tissueDir,"mo29112_ngs_dna_targdna_foundationone_20-2041_n65_one-sample-per-line_20200615.xlsx")
tissueOspl <- read_excel(ospl_tissue_file) %>%
  rename(SAMPLE.ID = "SAMPLE ID", SUBJECT.ID = "SUBJECT ID") %>% 
  mutate(SUBJECT.ID = as.character(SUBJECT.ID))

ospl = bind_rows(
  liqOspl %>% mutate( SAMPLE.ID = as.character(SAMPLE.ID) ),
  tissueOspl %>% mutate( Visit = "Tissue", SUBJECT.ID = as.character(SUBJECT.ID) )
  #  tissueOapl %>% select(SUBJECT.ID) %>% unique() %>% mutate(Visit = "Tissue")
) %>% left_join(
  patientData %>% mutate( SUBJECT.ID = as.character(SUBJECT.ID) )
) %>% 
  mutate(SUBJECT.ID = gsub("\\.0","",SUBJECT.ID))

# Stack tissue and liquid data
cycles = unique(ospl$Cycle)
cycles = cycles[ order( as.numeric( gsub("C(\\d+)D1[MI]P","\\1",cycles) ) ) ]

oapl <- bind_rows(
  tissueOapl %>% mutate(Visit = "Tissue"),
  liqOapl %>% mutate(SAMPLE.ID=as.character(SAMPLE.ID))
) %>%
  mutate(
    variant = gsub("-","",paste0(SV.PROTEIN.CHANGE,CNA.TYPE,REARR.DESCRIPTION)),
    baseEnd = ifelse(grepl("C1",Visit),"C1D1",
                     ifelse(grepl("PD",Visit),"PD",NA)),
    SUBJECT.ID = as.character(SUBJECT.ID),
    CycleN = as.numeric( gsub("C(\\d+)D1[MI]P","\\1",Cycle) ) - ifelse(grepl("C1D1IP",Cycle),1,0),
    Cycle = factor(Cycle,levels = cycles),
    SV.PERCENT.READS = as.numeric(SV.PERCENT.READS)
  ) %>%
  left_join(
    patientData
  )

msiScore <- oapl %>% filter(VARIANT.TYPE=="BIOMARKER-MSI SCORE")
msiStatus <- oapl %>% filter(VARIANT.TYPE=="BIOMARKER-MSI STATUS")
tmb <- oapl %>% filter(VARIANT.TYPE=="BIOMARKER-TMB")
oapl <- oapl %>% filter(!grepl("BIOMARKER",VARIANT.TYPE))

#re-extract tissue and liq
tissueOapl <- oapl %>% filter(Visit == "Tissue")
tissueOaplAll <- tissueOapl
tissueOapl <- tissueOapl %>% filter(SOMATIC.STATUS.FUNCTIONAL.IMPACT != "unknown")

liqOapl <- oapl %>% filter(Visit != "Tissue")
liqOaplAll <- liqOapl
liqOapl <- liqOapl %>% filter(SOMATIC.STATUS.FUNCTIONAL.IMPACT != "unknown")

oapl <- oapl %>% filter(SOMATIC.STATUS.FUNCTIONAL.IMPACT != "unknown")

pathwayGenes <- c("EGFR", "KRAS", "NRAS", "BRAF", "MAP2K1", "NF1")

```

MODUL (MO29112, NCT02291289) is an adaptable, phase 2, signal-seeking trial testing novel agents as first-line therapy for predefined subgroups of patients with metastatic colorectal cancer. It is comprised of four distinct cohorts. This analysis focuses on cohort 1, comprised of subjects with BRAF V600E mutant tumors, randomized to either the experimental arm of 5-FU/LV + cetuximab + vemurafenib or a control arm of FP + bevacizumab. Subjects' tumor tissue was tested with the FMI Foundation One CDx assay, and plasma samples from C1D1 and/or PD/EOT were tested with the FMI Foundation One Liquid CDx assay. Variants detected at these three points will be compared/contrasted to test hypotheses that treatment affected tumor genotype, specifically that cetuximab put a negative selective pressure on V600E mutations and a positive selective pressure on compensatory mutations elsewhere in the EGFR pathway. 

Here we plot the variant allele frequencies for FMI FOne Liquid samples. Patients with only one sample or no variants observed at multiple timepoints are removed.

```{r, fig.height = 13, fig.width = 9}
samplesPerPatient = liqOspl %>% group_by(SUBJECT.ID) %>% summarise(n = n())
multiSamplePatients = unlist( samplesPerPatient %>% filter( n > 1 ) %>% select(SUBJECT.ID) )

varsPerPatient = liqOapl %>% filter( !is.na(SV.PERCENT.READS) & !is.na(CycleN) ) %>% group_by(SUBJECT.ID,variant) %>% summarise(n = n())
multiVarPatients = unlist( varsPerPatient %>% filter( n > 1 ) %>% select(SUBJECT.ID) )

oapl2plot = liqOapl %>% filter(
  SUBJECT.ID %in% multiSamplePatients &
    SUBJECT.ID %in% multiVarPatients,
  !is.na(SV.PERCENT.READS)
) %>% 
  mutate(
    geneVariant = paste(GENE,variant)
  )

ggplot(oapl2plot, aes( x = CycleN, y = SV.PERCENT.READS, color = geneVariant )) +
  geom_point(cex=2) +
  geom_line(aes(group = geneVariant)) +
  facet_wrap(~SUBJECT.ID, ncol = 6, scales = "free") + 
  theme(legend.position="none")
```

And, bigger plots for patients with more sampling.

```{r, fig.height = 9, fig.width = 9}
samplesPerPatient = liqOspl %>% group_by(SUBJECT.ID) %>% summarise(n = n())
multiSamplePatients = unlist( samplesPerPatient %>% filter( n > 3 ) %>% select(SUBJECT.ID) )

varsPerPatient = liqOapl %>% filter( !is.na(SV.PERCENT.READS) & !is.na(CycleN) ) %>% group_by(SUBJECT.ID,variant) %>% summarise(n = n())
multiVarPatients = unlist( varsPerPatient %>% filter( n > 3 ) %>% select(SUBJECT.ID) )

oapl2plot = liqOapl %>% filter(
  SUBJECT.ID %in% multiSamplePatients &
    SUBJECT.ID %in% multiVarPatients,
  !is.na(SV.PERCENT.READS)
) %>% 
  mutate(
    geneVariant = paste(GENE,variant)
  )

ggplot(oapl2plot, aes( x = CycleN, y = SV.PERCENT.READS, color = geneVariant )) +
  geom_point(cex=2) +
  geom_line(aes(group = geneVariant)) +
  facet_wrap(~SUBJECT.ID, ncol = 2, scales = "free") + 
  theme(legend.position="none")
```
